Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/evaluation.tex
various edits, shrinking, and conclusion
adding number back in
submitted version

\newcommand\regname[1]{\texttt{#1}}
\newcommand\pointed{\regname{pointed}}
\newcommand\forward{\regname{forward}}

\noindent
To study the effectiveness of our approach, we applied our tool back
onto the corpus from \Cref{sec:study}.  We used the inference
procedure of the previous section to generate array specifications
for the corpus to assess the design of the language.
%
%We first examined how
%frequently stencil computations occur. We parsed
%\num{\overalllinesParsed} lines of Fortran code and found that
%\overalltickAssignPercent\% (\num{\overalltickAssign}) of statements
%have a left-hand side as an array subscript on neighbourhood
%indices. This supports the idea that stencil-like computations are
%common in scientific code.  
%
We did not expect to infer specifications for each array computation
we found because our analysis restricts the
array-subscript-statements, along the lines of~\Cref{sec:study}.
%A
%single statement can involve multiple arrays and therefore can have
%multiple specifications associated with it.
Overall we ended up with
\num{\overallnumStencilSpecs} specifications of which
\num{\overalltickLHSvar} were with a left-hand side variable rather
than a subscript (an \term{access} specification), leaving
\num{\overallStencilsSpecs} \term{stencil} specifications.
%
%For example, \num{\overalltickLHSvar}
%potential stencils were discarded because the left-hand side was not
%an array access. Overall, we were able to infer a stencil from
%\overalltickAssignSuccessPercentOfTickAssign\% of the statements
%involving an array assignment. A single statement can involve multiple
%arrays and.
This shows that we can express a large number of
array access patterns within our high-level abstractions and validates our
initial design choices informed by our data (~\Cref{subsec:choices}).

@@

\textbf{All pointed} \num{\overalljustPointed} of the specifications we
generated involved only \pointed\ regions.
% 39,681 of these were pointed in all dimensions.
Common examples of this were pointwise transformations on data (such
as scaling).

@@

@@

\noindent
There were various reasons why we did not infer
specifications on every looped array computation:
\begin{itemize}
\item \textbf{Computed induction variables} where an
index (\mintinline{fortran}{x}) is computed from an
induction variable via an intermediate definition, \eg{}, 
\mintinline{fortran}{x = i + 1}, or using functions that
we could analyse or classify;

@@

Detecting whether an error could have been prevented by a stencil
specification is difficult because it requires knowledge of the
programmer's intent. However, this is occasionally captured in commit
messages accompanying a code change. For example in the Unified Model
we identified an example of an off-by-1 error that could have been
avoided through the use of stencil specification. The buggy code was
identified as \pointed\ in all dimensions:

\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil readOnce, pointed(dim=1)*pointed(dim=2)*pointed(dim=3)
\end{minted}

But the programmer made it clear in the comments after the bug fix
that the intention had been \forward\ in the third dimension:

\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil readOnce, forward(depth=1, dim=3, nonpointed)*pointed(dim=1)*pointed(dim=2)
\end{minted}
%
We found a similar bug in 
Cliffs\footnote{\url{https://github.com/Delta-function/cliffs-src/commit/e2ca312cfc5398287c7ca8594e977658617c2540}}
where a subscript \fortran{u(i)} is corrected to \fortran{u(i-1)} which
could have been detected by an initial specification \fortran{backward(depth = 1,
  dim = 1, nonpointed)}.

>>>>

\newcommand{\overallStencilsSpecs}{77118}
\newcommand\regname[1]{\texttt{#1}}
\newcommand\pointed{\regname{pointed}}
\newcommand\forward{\regname{forward}}

\noindent
To study the effectiveness of our approach, we applied our tool back
onto the corpus from \Cref{sec:study}.  We used the inference
procedure of the previous section to generate array specifications
for the corpus to assess the design of the language.
%
%We first examined how
%frequently stencil computations occur. We parsed
%\num{\overalllinesParsed} lines of Fortran code and found that
%\overalltickAssignPercent\% (\num{\overalltickAssign}) of statements
%have a left-hand side as an array subscript on neighbourhood
%indices. This supports the idea that stencil-like computations are
%common in scientific code.
%
We did not expect to infer specifications for each array computation
we found because our analysis restricts the
array-subscript-statements, along the lines of~\Cref{sec:study}.
%A
%single statement can involve multiple arrays and therefore can have
%multiple specifications associated with it.
Overall we ended up with
\num{\overallnumStencilSpecs} specifications of which
\num{\overalltickLHSvar} were with a left-hand side variable rather
than a subscript (an \term{access} specification), leaving
\num{\overallStencilsSpecs} \term{stencil} specifications.
%
%For example, \num{\overalltickLHSvar}
%potential stencils were discarded because the left-hand side was not
%an array access. Overall, we were able to infer a stencil from
%\overalltickAssignSuccessPercentOfTickAssign\% of the statements
%involving an array assignment. A single statement can involve multiple
%arrays and.
This shows that we can express a large number of
array access patterns within our high-level abstractions and it validates our
initial design choices informed by our data (\Cref{subsec:choices}).

@@

\textbf{All pointed} \num{\overalljustPointed} of the specifications we
generated involved only \pointed\ regions.
% 39,681 of these were pointed in all dimensions.
Common examples of this were pointwise transformations on data (such
as scaling). The remaining \num{\overallnonTrivial} we consider to
be non-trivial, \ie{} with a higher potential for programming error.

@@

\textbf{ReadOnce} specification occurred \num{\overallreadOnce} times, which
correlates with the high proportion of unique-subscript expressions
in the initial study.

@@

\noindent
There were various reasons why we did not infer
specifications on every looped array computation:
\begin{itemize}
\item \textbf{Computed induction variables} where an
index (\mintinline{fortran}{x}) is computed from an
induction variable via an intermediate definition, \eg{}, 
\mintinline{fortran}{x = i + 1}, or using functions that
we could not analyse or classify;

@@

Detecting whether an error could have been prevented by a stencil
specification is difficult because it requires knowledge of the
programmer's intent. However, this is occasionally captured in commit
messages accompanying a code change. For example in the Unified Model
we identified an example of an off-by-1 error that could have been
avoided through the use of stencil specification. The buggy code was
identified as \pointed\ in all dimensions by our inference:

\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil readOnce, pointed(dim=1)*pointed(dim=2)*pointed(dim=3)
\end{minted}
%
But the programmer made it clear in the comments after the bug fix
that the intention had been \forward\ in the third dimension, \ie{}, a
specification:
%
\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil readOnce, forward(depth=1, dim=3, nonpointed)*pointed(dim=1)*pointed(dim=2)
\end{minted}
%
We found a similar bug in 
Cliffs\footnote{\url{https://github.com/Delta-function/cliffs-src/commit/e2ca312cfc5398287c7ca8594e977658617c2540}}
where a subscript \fortran{u(i)} is corrected to \fortran{u(i-1)} which
could have been detected by an initial specification of \fortran{!=
  stencil backward(depth=1, dim=1, nonpointed)}.
